---
output:
  BiocStyle::html_document
---

<!--
%% \VignetteEngine{knitr::rmarkdown}
-->

---
title: "doppelgangR vignette"
author: "Levi Waldron and Markus Riester"
output:
  BiocStyle::html_document:
    toc: true
---



```{r, echo=FALSE, message=FALSE}
library(doppelgangR)
```

# Introduction
\Biocpkg{doppelgangR} is a package for identifying duplicate samples
within or between datasets.

## The importance of batch correction

We load for datasets by Yoshihara \emph{et al.} that have been curated
in \Biocpkg{curatedOvarianData}:

```{r}
library(curatedOvarianData)
data(GSE32062.GPL6480_eset)
data(GSE17260_eset)
```

The \Rfunction{doppelgangR} function requires a list of
\Rclass{ExpressionSet} objects as input, which we create here:

```{r testesets}
testesets <- list(JapaneseA=GSE32062.GPL6480_eset,
                  Yoshihara2010=GSE17260_eset)
```

We curate the alt\_sample\_name metadata simply by removing non-digits,
which then can be matched across studies:

```{r curateids}
testesets <- lapply(testesets, function(X){
    # standardize the sample ids to improve matching based on clinical annotation
    sampleNames(X) <- make.names(paste(X$sample_type,
        gsub("\\D","",X$alt_sample_name), sep="_"))
    X$alt_sample_name <- sampleNames(X)
    pData(X) <- pData(X)[, !grepl("uncurated_author_metadata", colnames(pData(X)))]
    X })
```

Run doppelgangR with default arguments:
```{r rundopp, results="hide", message=FALSE}
results1 <- doppelgangR(testesets, phenoFinder.args=NULL, cache.dir=NULL)
```

This creates an object of class \Rclass{DoppelGang}, which has print, summary, and plot methods.  Summary method output not shown here due to voluminous output:
```{r summarizedop, results='hide'}
summary(results1)
```

Plot creates a histogram of sample pairwise correlations within and between each study:

```{r plotdop, fig.cap="Doppelgangers identified on the basis of similar expression profiles.  The vertical red lines indicate samples that were flagged."}
par(mfrow=c(2,2), las=1)
plot(results1)
```

To create a data.frame of potential doppelgangers that you can write
to file then open in a spreadsheet to examine in detail, use the
summary method:
```{r }
res.df <- summary(results1)
dim(res.df)
```

To illustrate the importance of ComBat batch correction, we run again without batch correction:

```{r rundop2, results="hide", message=FALSE}
results2 <- doppelgangR(testesets, corFinder.args=list(use.ComBat=FALSE), phenoFinder.args=NULL, cache.dir=NULL)
```

Now only 4 (instead of 43) doppelganger pairs are found:
```{r summarizedop2}
dim(summary(results2))
```

We define a couple functions to assess which is more accurate, with or
without batch correction.  The first function analyzes two datasets
from a DoppelGang object: Iterating through each sample in the larger
dataset, it selects the sample in the smaller dataset with highest
correlation (doppelmelt).  This potential match is considered a True
Positive if the sample name is the same.

```{r output2rocfunc}
doppelmelt <- function(obj, ds1, ds2){
    if(paste(ds1, ds2, sep=":") %in% names(obj@fullresults)){
        ds <- paste(ds1, ds2, sep=":")
    }else if(paste(ds2, ds1, sep=":") %in% names(obj@fullresults)){
        ds <- paste(ds2, ds1, sep=":")
    }else{
        return(NULL)
    }
    cormat <- obj@fullresults[[ds]]$correlations
    if(nrow(cormat) < ncol(cormat)) cormat <- t(cormat)
    idx <- sapply(rownames(cormat), function(x) which.max(cormat[x, ]))
    corvec <- sapply(1:nrow(cormat), function(i) cormat[i, idx[i]])
    output <- data.frame(sample1=rownames(cormat),
                         sample2=colnames(cormat)[idx],
                         cor=corvec, stringsAsFactors=FALSE)
    output$truepos <- sub(".+:", "", output[, 1]) == sub(".+:", "", output[, 2])
    return(output)
}
plotROC <- function(pred, labels, plot = TRUE, na.rm = TRUE, colorize = FALSE, addtext=TRUE, ...) {
    require(ROCR)
    require(pROC)
    if (na.rm) {
        idx <- !is.na(labels)
        pred <- pred[idx]
        labels <- labels[idx]
    }
    pred.rocr <- prediction(pred, labels)
    perf.rocr <- performance(pred.rocr, "tpr", "fpr")
    auc <- performance(pred.rocr, "auc")@y.values[[1]][[1]]
    roc.obj <- roc(labels, pred)
    auc.ci <- ci(roc.obj)
    significant <- ifelse(ci(roc.obj, conf.level=0.9)[1] > 0.5, "*", "")
    best <- coords(roc.obj,x="best")
    if (plot) {
        plot(perf.rocr, colorize = colorize, cex.lab = 1.3, ...)
        abline(a = 0, b = 1, lty = 2)
        if(addtext){
        text(0, 0.9, paste("AUC = ", round(auc, digits = 2), significant,
                           sep=""), cex = 1.5, pos = 4)
        text(1, 0.1, paste("n =", length(labels)), cex = 1.5, pos = 2)
    }
    }
    invisible(list(auc,auc.ci,best))
}
```

```{r yoshihararoc, fig.height=3, fig.cap="Identification of Duplicate samples between the Japanese A dataset and Yoshihara 2010 dataset, without (left) and with (right) batch correction."}
roc1 <- doppelmelt(results1, "JapaneseA", "Yoshihara2010")
roc2 <- doppelmelt(results2, "JapaneseA", "Yoshihara2010")
par(mfrow=c(1,2), las=1)
plotROC(roc2$cor, roc2$truepos, main="No batch correction")
plotROC(roc1$cor, roc1$truepos, main="Batch correction")
```


%% rocobj.cv <- ROC(test=preds.cv,
%%     PV=FALSE,MX=FALSE,MI=FALSE,AUC=FALSE,
%%     stat=(thisstat==levels(thisstat)[2]),
%%     plot="ROC")
%% rocobj.resubstitution <- ROC(test=preds.resubstitution,
%%     PV=FALSE,MX=FALSE,MI=FALSE,AUC=FALSE,
%%     stat=(thisstat==levels(thisstat)[2]),
%%     plot="none")
%% with(rocobj.resubstitution$res,lines(1-spec,sens),lty=3)
%% legend("bottomright",lty=1,pch=-1,lw=1:2,
%%        legend=c(paste("RESUBSTITUTION area under curve =",round(rocobj.resubstitution$AUC,2)),
%%          paste("CROSS-VALIDATED area under curve =",round(rocobj.cv$AUC,2))))

\section{The impact of duplicates on prognostic model validation}

```{r <>}
library(simulatorZ)
library(survival)
esets <- testesets
for (i in 1:length(esets)){
    esets[[i]]$y <- Surv(esets[[i]]$days_to_death, esets[[i]]$vital_status=="deceased")
}
##survmodel <- rowCoxTests(X=esets[["JapaneseA"]], y="y")  #doesn't work
modelcoefs <- rowCoxTests(X=exprs(esets[["Yoshihara2010"]]), y=esets[["Yoshihara2010"]]$y)
##modelcoefs$coef[modelcoefs$p.value > 0.1] <- 0
preds <- data.frame(y=esets[["JapaneseA"]]$y,
                    linpred=t(exprs(esets[["JapaneseA"]])) %*% sign(modelcoefs$coef),
                    is.dup=sampleNames(esets[["JapaneseA"]]) %in% sampleNames(esets[["Yoshihara2010"]]))
preds <- preds[order(preds$is.dup, decreasing=TRUE), ]
max.dup <- max(which(preds$is.dup))
preds$percent.dups <- sapply(1:nrow(preds), function(i){
    if(i > max.dup) return(0)
    return( (max.dup - i + 1) / nrow(preds) * 100)
})

hr.removedups <- t(sapply(0:max.dup, function(i){
    if(i>0){
        dat <- preds[-1:-i, ]
    }else{
        dat <- preds
    }
    coxfit <- coxph(y ~ (linpred > median(linpred)), data=dat)
    output <- (summary(coxfit))$conf.int
    names(output) <- c("exp(coef)", "exp(-coef)", "lower .95", "upper .95")
    output
}))
hr.removedups <- data.frame(hr.removedups)
hr.removedups$ndups <- (nrow(hr.removedups)-1):0

hr.removedups$percent.dups <- preds$percent.dups[1:nrow(hr.removedups)]
midlowess <- lowess(exp.coef. ~ percent.dups, data=hr.removedups)
##lowerlowess <- lowess(lower..95 ~ ndups, data=hr.removedups)
##upperlowess <- lowess(upper..95 ~ ndups, data=hr.removedups)
```

HR vs. percent duplication plot:

```{r }
par(las=1)
plot(exp.coef. ~ percent.dups, data=hr.removedups, ylab="Hazard Ratio", xlab="% of samples duplicated in training set")
lines(midlowess$x, midlowess$y, lw=2)
```

\section{Finding duplicates between microarray and RNA-seq data}

Load TCGA microarray and RNA-seq data for ovarian cancer.  These are
the level III RNA-seq data, as provided by the Data Portal (summarized
at gene level).  Note that these are the same patients, profiled by
microarray and RNA-seq.  We would like to see how well we can identify
the duplicates based on expression data alone.

```{r tcgaovcload}
library(curatedOvarianData)
data(TCGA.RNASeqV2_eset)
data(TCGA_eset)
tcgaov.esets <- list(microarray=TCGA_eset, rnaseq=TCGA.RNASeqV2_eset)
```

```{r rundoptcga, results="hide", message=FALSE}
tcgaout.batch <- doppelgangR(tcgaov.esets, phenoFinder.args=NULL, smokingGunFinder=NULL, cache.dir=NULL)
tcgaout.nobatch <- doppelgangR(tcgaov.esets, phenoFinder.args=NULL, smokingGunFinder=NULL,
                               corFinder.args=list(use.ComBat=FALSE), cache.dir=NULL)
```

```{r plottcta, fig.height=4.7, fig.caption="Microarray samples identified as having RNA-seq duplicates in the TCGA ovarian cancer dataset."}
par(mfrow=c(2, 3), las=1)
plot(tcgaout.batch, xlim=c(0.6, 1))
plot(tcgaout.nobatch, xlim=c(0.6, 1))
```

```{r tcgaroc, fig.height=3.5, fig.cap="Identification of Affymetrix microarray profiles that have duplicate RNA-seq profiles in the TCGA ovarian cancer dataset.  Note that most of the errors after batch correction are due to an RNA mix-up in the TCGA Affymetrix dataset."}
tcgamelt.batch <- doppelmelt(tcgaout.batch, "microarray", "rnaseq")
tcgamelt.nobatch <- doppelmelt(tcgaout.nobatch, "microarray", "rnaseq")
tcga.remove <- read.delim(system.file("extdata", "TCGA_remove.txt",
                    package="doppelgangR"), as.is=TRUE)[, 1]
par(mfrow=c(1,3), las=1)

nobatch.rocr <- with(tcgamelt.nobatch, prediction(cor, as.integer(truepos)))
withbatch.rocr <- with(tcgamelt.batch, prediction(cor, as.integer(truepos)))
remove.rocr <- with(tcgamelt.batch[!sub("microarray:", "", tcgamelt.batch[, 1]) %in% tcga.remove, ],
                    prediction(cor, as.integer(truepos)))
```

Plotting of three ROCs in one for RNA-seq vs microarray in TCGA
ovarian cancer.

```{r, echo=FALSE}
#pdf("TCGA-rnaseq.pdf", width=3, height=3)
#par(mar=c(4, 4, 0.2, 0.2))
#dev.off()
```

```{r}
par(las=1)
nobatch.auc <- round(performance(nobatch.rocr, "auc")@y.values[[1]][[1]], 2)
withbatch.auc <- round(performance(withbatch.rocr, "auc")@y.values[[1]][[1]], 2)
remove.auc <- round(performance(remove.rocr, "auc")@y.values[[1]][[1]], 2)
plot(performance(nobatch.rocr, "tpr", "fpr"), lty=3)
plot(performance(withbatch.rocr, "tpr", "fpr"), add=TRUE, type="l", , lty=2)
plot(performance(remove.rocr, "tpr", "fpr"), add=TRUE, type="l", lty=1, lw=2)
legend("bottomright", lty=1:3, bty="n", cex=0.7,
       legend=c(paste("Mix-ups removed (AUC=", remove.auc, ")", sep=""),
           paste("Batch correction (AUC=", withbatch.auc, ")", sep=""),
           paste("No correction (AUC=", nobatch.auc, ")", sep="")))
```

\section{Application to breast cancer}

To save time in building the vignette, six breast cancer datasets
available as Bioconductor packages are installed, loaded, and prepared
\emph{outside} of this vignette.  The following code for preparing the ExpressionSets is not run:

```{r createbreastesets, eval=FALSE}
breast.packages <- c("breastCancerMAINZ", "breastCancerNKI",
    "breastCancerTRANSBIG", "breastCancerUNT", "breastCancerUPP",
    "breastCancerVDX")
##
other.packages <- "WGCNA"
##
if (!require(BiocInstaller))
    stop("You need to install Bioconductor, which includes BiocInstaller.")
##
for (pkg in breast.packages){
    if(!require(package=pkg, character.only=TRUE)){
        print(paste("Need to install", pkg))
        biocLite(pkg, suppressUpdates=TRUE, suppressAutoUpdate=TRUE, ask=FALSE)
    }
}
##
for (pkg in other.packages){
    if(!require(package=pkg, character.only=TRUE)){
        print(paste("Need to install", pkg))
        biocLite(pkg, suppressUpdates=TRUE, suppressAutoUpdate=TRUE, ask=FALSE)
    }
}
##
esets <- lapply(breast.packages, function(pkg){
    print(pkg)
    library(Biobase)
    esetname <- tolower(sub("breastCancer", "", pkg))
    data(list=esetname)
    output <- get(esetname)
    output <- output[!is.na(featureData(output)$EntrezGene.ID), ]
    merge.probeset <- WGCNA::collapseRows(datET=exprs(output),
                                          rowGroup=featureData(output)$EntrezGene.ID,
                                          rowID=featureNames(output))
    output <- output[merge.probeset$selectedRow, ]
    featureNames(output) <- featureData(output)$EntrezGene.ID
    return(output)
})
names(esets) <- sub("breastCancer", "", breast.packages)
##
save(esets, file="esets_breast.rda")
```

Here we just load the prepared ExpressionSets.  These have had
probesets collapsed to Entrez Gene identifiers, using the
WGCNA::collapseRows() function using the probeset with maximum mean to
represent each gene.
```{r loadbreastesets}
load(url("http://bcb.dfci.harvard.edu/ovariancancer/dfiles_old/esets_breast.rda"))
```

Now run doppelgangR for UNT + UPP only.
```{r breastdoppel}
output <- doppelgangR(esets[c("UNT", "UPP")], outlierFinder.expr.args = list(bonf.pvalue=10, transFun=atanh, tail="upper"),
                      outlierFinder.pheno.args=NULL, cache.dir=NULL)
output
```

```{r plotbreastdoppel}
par(mfrow=c(2,2), las=1)
plot(output)
```

%% \section{Impact of doppelgangers on model validation}
%% Hidden doppelgangers present in both a training and test set can substantially inflate estimates of prediction model accuracy.  We demonstrate this using the UNT and UPP breast cancer datasets.

%% ```{r }
%% eset.training <- esets["UPP"]
%% eset.test <- esets["UNT"]
%% ```

\section{Cell-line data}

Download pre-processed Affymetrix expression data for all NCI60 and CCLE:

```{r cell_line_load}
download.path <- "."

download.file("http://s3.amazonaws.com/cancerhub/celllines/NCI60_mrna.rds", file.path(download.path, "NCI60_mrna.rds"))
download.file("http://s3.amazonaws.com/cancerhub/celllines/CCLE_mrna.rds", file.path(download.path, "CCLE_mrna.rds"))
nci60.eset <- readRDS(file.path(download.path, "NCI60_mrna.rds"))
ccle.eset <- readRDS(file.path(download.path, "CCLE_mrna.rds"))
nci60.eset$tissue <- gsub("_.*$","", sampleNames(nci60.eset))
ccle.eset$tissue <- gsub("^.*_","", sampleNames(ccle.eset))
eset.list <- list(CCLE=ccle.eset, NCI60=nci60.eset)
````

Run doppelgangR without trying to match clinical annotation:
```{r cell_line_run}
ret <- doppelgangR(eset.list, phenoFinder.args=NULL, smokingGunFinder.args=NULL, outlierFinder.expr.args=list(bonf.prob = 3.0, transFun = atanh, tail = "upper"))
```

Now evaluate the results, using a curated mapping NCI60 to CCLE sample id mapping file:
```{r cell_line_evaluate}
nci60_ccle <- read.csv(system.file("extdata", "NCI60_CCLE_overlap.csv",
                    package="doppelgangR"), as.is=TRUE)

roc1 <- doppelmelt(ret, "NCI60", "CCLE")

true.match <- gsub("NCI60:","",roc1[,2])== nci60_ccle$NCI60[match(gsub("CCLE:","",roc1[,1]), nci60_ccle$CCLE)]
table(true.match)
roc1$truepos[!is.na(true.match)] <- true.match[!is.na(true.match)]
plotROC(roc1$cor, roc1$truepos, main="NCI60/CCLE doppelgangers")
```

\section{Parallelization}
DoppelgangR checks for duplicates within each dataset in the list of
ExpressionSets, and between each pair of datasets.  This is an
``embarassingly parallel'' process that can be divided among multiple
processors.  Parallelization is implemented via

\end{document}
