---
title: "doppelgängR"
author: "Levi Waldron"
date: "`r format(Sys.time(), '%B %d, %Y')`"
package: "`r pkg_ver('doppelgangR')`"
vignette: >
  %\VignetteIndexEntry{doppelgangR vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: >
    Whole-genome analysis of cancer specimens is commonplace and
    investigators frequently share or re-use specimens in later studies.
    Duplicate expression profiles in public databases will impact
    re-analysis if left undetected, a so-called “doppelgänger” effect. The
    doppelgängR package uses batch correction and outlier detection among
    pairwise expression profile correlations to accurately identify
    duplicate profiles for cancer types where profiles are sufficiently
    distinct.  It is intended for use when nucleotide-level sequence data
    are unavailable, and is is effective even for specimens where
    duplicated samples are profiled by different microarray technologies,
    or by a combination of microarray and log-transformed RNA-seq data.
output:
  BiocStyle::html_document:
  toc: true
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
opts_knit$set(cache=TRUE)
has_data <- requireNamespace("curatedOvarianData", quietly = TRUE)
opts_chunk$set(eval = has_data)
```

# doppelgängR

## Introduction

`r BiocStyle::Biocpkg("doppelgangR")` is a package for identifying duplicate
samples within or between datasets of transcriptome profiles. It is intended for
microarray and RNA-seq gene expression profiles where biological replicates are
ordinarily more distinct than technical replicates, as is the case for cancer
types with "noisy" genomes. It is intended for cases where per-gene summaries
are available but full genotypes are not, which is typical of public databases
such as the Gene Expression Omnibus.

Results from running `doppelgangR` on CRC, bladder, and ovarian
[on Dropbox](httpss://www.dropbox.com/sh/or3l7oo96x1ilm8/AAD1MVQVnoS6H7OExoYwOovSa?dl=0).

For the manuscript vignette, visit
<https://github.com/waldronlab/doppelgangR_paper>.

```{r loadpkg, echo=FALSE, message=FALSE}
library(doppelgangR)
```

The `doppelgangR()` function identifies duplicates in three different ways:

* **"expression"** doppelgängers have highly similar expression profiles, which
are identified by default by having higher Pearson correlation than expected
based on an empirical distribution of Pearson correlations between biological
replicates. The type of correlation, and default use of ComBat batch correction,
can be changed using the "corFinder.args" argument.

* **"phenotype"** doppelgängers have highly similar clinical or phenotype data,
as contained in the phenoData slot of the `ExpressionSet`. In order to identify
duplicates this way, it is required to curate the phenoData of each
ExpressionSet they have identical column names, and encode phenotypes in the
same way.  For example, if each dataset provides information on age, this column
of the phenoData could be called "age" in every dataset, and encoded as an
integer number of years.  If the phenoData slots are NULL then this type of
checking will automatically be turned off.  If they are not NULL but are also
not curated, you should turn off phenotype checking by setting
`phenoFinder.args=NULL`.

* **"smoking gun"** doppelgängers have the same value for an identifier that
should be unique. You can enable this type of check by setting the argument
"manual.smokingguns" to the names of columns containing supposedly unique
identifiers, or setting "automatic.smokingguns" to TRUE, and the function will
assume any column containing unique values within the column should also be
unique across datasets.

This vignette focuses on the "expression" type of doppelgänger.

# Data types

Identification of doppelgängers is effective for both microarray and
**log-transformed** RNA-seq data, and even for matching samples that have been
profiled by microarray and RNA-seq.

# Case Study: Batch correction in Japanese datasets

We load for datasets by Yoshihara __et al.__ that have been curated
in `r BiocStyle::Biocexptpkg("curatedOvarianData")`.  These are objects of class `ExpressionSet`.

```{r loaddata, message = FALSE}
library(curatedOvarianData)
data(GSE32062.GPL6480_eset)
data(GSE17260_eset)
```

The `doppelgangR` function requires a list of
`ExpressionSet` objects as input, which we create here:

```{r testesets}
testesets <- list(JapaneseA=GSE32062.GPL6480_eset,
                  Yoshihara2010=GSE17260_eset)
```

Now run `doppelgangR` with default arguments, except for setting
`phenoFinder.args=NULL`, which turns off checking for similar clinical data in
the `phenoData` slot of the ExpressionSet objects:

```{r rundopp, results="hide", message=FALSE, cache=TRUE}
results1 <- doppelgangR(testesets, phenoFinder.args=NULL)
```

This creates an object of class `DoppelGang`, which has print, summary, and plot
methods.  Summary method output not shown here due to voluminous output:

```{r summarizedop, results='hide'}
summary(results1)
```

Plot creates a histogram of sample pairwise correlations within and between each
study:

```{r plotdop}
#| fig-cap: Doppelgängers identified on the basis of similar expression
#|   profiles. The vertical red lines indicate samples that were flagged.

par(mfrow=c(2,2), las=1)
plot(results1)
```

One of these histograms can be drawn using the plot.pair argument:

```{r plotdop2}
#| fig-cap: Pair plot of JapaneseA:JapaneseA Doppelgängers identified. The
#|   vertical red lines indicate samples that were flagged.

plot(results1, plot.pair=c("JapaneseA", "JapaneseA"))
```

# Important options

## Changing sensitivity

If after inspecting the histograms, you see that some visible outliers were not
caught, or non-outliers exceeded the sensitivity threshold, you can change the
default sensitivity using the argument:

```{r changesens, eval=FALSE}
outlierFinder.expr.args =
    list(bonf.prob = 0.5, transFun = atanh, tail = "upper")
```

The default 0.5 is a reasonable but arbitrary trade-off between sensitivity and
specificity which we have found to often select dataset pairs containing
duplicates, but to often not find *all* the duplicate samples.  Sensitivity can
be increased by changing the bonf.prob argument, *i.e.*:

```{r rundopp2, results="hide", message=FALSE, eval=FALSE}
results1 <- doppelgangR(testesets, 
        outlierFinder.expr.args = list(bonf.prob = 1.0, transFun = atanh, 
                                       tail = "upper"))
```

## Use of the ExpressionSet

The `doppelgangR()` function takes as its main argument a list of
`ExpressionSet` objects.  If you just have matrices, you can easily convert
these to the `ExpressionSet` objects, for example:

```{r makematrix}
mat <- matrix(1:4, ncol=2)
library(Biobase)
eset <- ExpressionSet(mat)
class(eset)
```

## Parallelizing

The `doppelgangR()` function checks all pairwise combinations of datasets in a
list of `ExpressionSet` objects, and these dataset pairs can be checked in
parallel using multiple processing cores. This functionality is imported from
the "future" package. Please see the "future" package documentation to configure
your parallel backend.

```{r parallel, eval=FALSE}
library(future)
plan(multisession, workers = 8)
results2 <- doppelgangR(testesets)
```

## Caching

By default, the `doppelgangR()` function caches intermediate results to make
re-running with different arguments faster.  Turn caching off by setting the
argument `cache.dir=NULL`.

# Session Info

```{r sessionInfo, eval=TRUE}
sessionInfo()
```
